p-adic stability in linear algebra

### 2015-01-13 examples/stability.tex
Working on characteristic polynomial section

\newtheorem{theo}{Theorem}[section]
\newtheorem{lem}[theo]{Lemma}
\newtheorem{prop}[theo]{Proposition}
\newtheorem{cor}[theo]{Corollary}
\newtheorem{quest}[theo]{Question}
%\theoremstyle{definition}
\newtheorem{rem}[theo]{Remark}
\newtheorem{ex}[theo]{Example}
\newtheorem{deftn}[theo]{Definition}
\newtheorem{rmk}[theo]{Remark}

@@

\newcommand{\low}{\text{\rm low}}
\newcommand{\up}{\text{\rm up}}

\newcommand{\detp}{\text{\rm det}'}
\newcommand{\DI}{\text{\rm DI}}
\newcommand{\II}{\text{\rm II}}

@@

\numberofauthors{3}
\author{
\alignauthor Xavier Caruso\\
  \affaddr{Universit\'e Rennes 1}\\
  \email{\normalsize \textsf{xavier.caruso@normalesup.org}}
\alignauthor Tristan Vaccon\\
  \affaddr{Universit\'e Rennes 1}\\
  \email{\normalsize \textsf{tristan.vaccon@univ-rennes1.fr}}
\alignauthor David Roe \\
  \affaddr{University of Calgary}\\
  \email{\normalsize \textsf{roed.math@gmail.com}}
}

@@

\begin{abstract}
Using the differential precision methods developed previously by the same authors,
we study the $p$-adic stability of standard operations on matrices and vector 
spaces. We demonstrate that lattice-based methods surpass naive methods in many
applications, such as matrix multiplication and sums and intersections of subspaces.
We also analyze determinants, characteristic polynomials and LU factorization using these differential methods.
We supplement our observations with numerical experiments.
\end{abstract}

@@

To make this sentence more explicit, let us consider a finite 
dimensional\footnote{The framework of \cite{caruso-roe-vaccon:14a} is 
actually those of Banach spaces. However we will not need infinite 
dimensional spaces in the present paper.} normed vector space $E$ 
defined over $K$. We use the notation $\Vert \cdot \Vert_E$ for the norm 
on $E$ and $B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. 
closed) ball of radius $r$ centered at the origin. A lattice $L \subset 
E$ is, by definition, a sub-$\O_K$-module which generates $E$ over $K$. 
Since we are working in a ultrametric world, the balls $B^{\phantom 
-}_E(r)$ and $B^-_E(r)$ are examples of lattices. Actually, lattices 
should be thought as special neighborhoods of $0$ and therefore are good 
candidates to model precision data. Moreover, as revealed in 
\cite{caruso-roe-vaccon:14a}, they behave quite well under (strictly) 
differentiable maps:

@@

\begin{rem}
In certain situations,
it may happen that the function $f$ is solution of a simplier 
differential equation of the form $f' = g \circ f$. If this holds, 
Proposition \ref{prop:boundLambdaf2} gives the bound $\Lambda(f)_{\leq 
2} (x) \leq 2(x + \alpha + \mu) - \nu$ for $x \leq \nu - \mu - \alpha$.

Beyond this particular case, our general recommandation for choosing the 
function $h$ is to endow $F$ with the second norm $\Vert x \Vert'_F = 
e^\mu \cdot \Vert x \Vert_F$ ($x \in F$) and to take $h : (F, \Vert 
\cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ acting as identity on the
underlying vector spaces. The function $\Lambda(h)$ the maps $x$ to $x + 
\mu$ ($x \in \R$) and we can choose $\Lambda_h = \Lambda(h)$.
\end{rem}

@@

@@

We now come to the computation of the determinant of the square $n 
\times n$ with coefficients in $K$. The differential computation is 
quite classical: the differential of the function $\det : M_{n,n}(K) \to 
K$ at some point $M$ is the linear mapping:
$$\detp(M) : dM \mapsto \tr(\com(M) \cdot dM)$$
where $\com(M)$ stands for the comatrix of $M$. If $M$ is not 
invertible, $\detp(M)$ is not surjective and we cannot apply Proposition 
\ref{prop:precision}. Therefore, from now on, we suppose that $M$ is invertible.

@@

In terms of precision, Proposition \ref{prop:detmatrix} implies that if 
$M$ is given at flat precision $O(\pi^N)$ with $N > v$, then $\det(M)$ 
is known at precision $O(\pi^{N+v})$, so we are gaining $v$ significant 
digits in absolute precision\footnote{We note that the valuation of 
$\det(M)$ is $v + m_n$; thus if we are considering \emph{relative} 
precisions then we should say that we are loosing $m_n$ significant 
digits.}.
Furthermore, there exists a easy method for computing the determinant
of $M$ with this optimal precision. It consists in first computing an
approximate Smith decomposition of $M$, that is a writing
$M = \tilde U_M \cdot \tilde \Delta_M \cdot \tilde V_M$
where $\tilde U_M$ and $\tilde V_M$ are exact unimodular matrices and
$\tilde \Delta_M$ is known at precision $O(\pi^N)$, is diagonal modulo
$\pi^N$ and has diagonal entries $a_1, \ldots, a_m$ of valuations 
$m_1, \ldots, m_n$ respectively.
We then have $\det(M) = \det(\tilde \Delta_M)$ and factoring $a_i$ on
the $i$-th row of $\tilde \Delta_M$, we obtain
$\det(\tilde \Delta_M) = a_1 \cdots a_m \cdot U$
where $U$ is a matrix which is congruent to the identity modulo 
$\pi^{N-m_n}$. Therefore we get:
$$\det(M) \equiv a_1 \cdots a_m \pmod{\pi^{N+v}}$$
from what we see that we can compute $\det(M)$ at the required precision
by simply evaluating the product $a_1 \cdots a_m$.

@@

\todo{Write this subsection}

@@

In this subsection, we study the LU factorization.
%This example has been previously studied by Caruso in \cite{caruso:12a}, 
%but without using differential precision. We are going to show here that 
%the framework of \S \ref{ssec:boundLambdaf} applies directly.
We first introduce several notations.
We denote by $\Vert \cdot \Vert$ the subordinate matrix norm on $M_n(K)$
and, given a positive real number $C$, we let $B(C)$ be the closed ball 
in $M_n(K)$ centered at the origin of radius $C$. We consider in addition
the following subsets of $M_n(K)$:

\noindent $\bullet$
$O_n$ is the open subset of matrices whose all principal 
minors do not vanish (we recall that the latest condition implies the
existence and the unicity of a LU factorization);

\noindent $\bullet$
$U_n$ is the sub-vector space of upper-triangular 
matrices;

\noindent $\bullet$
$L_n$ (resp. $L_n^u$) is the sub-affine space space of 
nilpotent (resp. unipotent) lower-triangular matrices.

\noindent
{\bf Calculus and precision.}
We choose to normalize the LU factorization by requiring that $L$ is 
unipotent and denote by $D : O_n \to L_n^u \times U_n$ the function 
modeling this decomposition. The computation of the differential of $D$ 
has already been done in \cite[Appendix B]{caruso-roe-vaccon:14a}. For
$M \in \O_n$ with $D(M) = (L,U)$, the linear mapping $D'(M)$ is given 
by:
$$\begin{array}{l}
dM \mapsto (L \cdot \low(dX), \up(dX) \cdot U) \smallskip \\
\hspace{1cm}\text{with } dX = L^{-1} \cdot dM \cdot U^{-1}
\end{array}$$
where $\low$ (resp. $\up$) denotes the canonical projection of $M_n(K)$ 
onto $L_n$ (resp. $U_n$). 
It is easily checked that $D'(M)$ is bijective with inverse given by
$(A,B) \mapsto AU+LB$.

We now want to apply Proposition \ref{prop:boundLambdaf2}
in order to derive a concrete result on precision. We pick $M_0 \in
O_n$ and write $D(M_0) = (L_0,U_0)$. We consider the ``translated'' 
function $f$ taking $M$ to $D(M_0+M)-D(M_0)$. We then have $f(0) = 0$ 
and $f'(M) = D'(M_0+M)$. Using the explicit description of the inverse
of $D'(M_0+M)$, we find $B(1) \subset f'(0) \cdot B(C)$
for $C = \max(\Vert U_0 \Vert, \Vert L_0 \Vert)$. Moreover $f$ satisfies 
the differential equation $f' = g \circ f$ where $g$ is defined by:
\begin{equation}
\label{eq:gAB}
\begin{array}{l}
g(A,B)(X) = \big( (L_0+A) \cdot \low(Y), \up(Y) \cdot (U_0+B) \big)
\smallskip \\
\hspace{1.8cm}\text{with } Y = (L_0+A)^{-1} \cdot X \cdot (U_0+B)^{-1}.
\end{array}
\end{equation}
Let $\kappa(S) = \Vert S \Vert \cdot \Vert S^{-1} \Vert$ denote the 
condition number of a matrix $S$. Remarking that $\Vert S + T 
\Vert = \Vert S \Vert$ if $\Vert T \Vert < \Vert S \Vert$ and $\Vert (S 
+ T)^{-1} \Vert = \Vert S^{-1} \Vert$ if $\Vert T \Vert < \Vert S^{-1} 
\Vert^{-1}$, we deduce from \eqref{eq:gAB} that
$\Vert g(A,B) \Vert \leq \max 
\big( \kappa(L_0) \Vert U_0^{-1} \Vert, 
\kappa(U_0) \Vert L_0^{-1} \Vert \big)$
provided that $\Vert A \Vert < \Vert L_0^{-1} \Vert^{-1}$ and 
$\Vert B \Vert < \Vert U_0^{-1} \Vert^{-1}$.
Combining this with Proposition \ref{prop:boundLambdaf2} and 
\cite[Proposition.~3.12]{caruso-roe-vaccon:14a}, we finally find that 
Eq.~\eqref{eq:firstorder} holds as soon as 
$$\begin{array}{l}
\displaystyle \frac \rho r > \Vert p \Vert^{-\frac{2p}{p-1}} \cdot
\max ( \Vert L_0 \Vert, \Vert U_0 \Vert) \cdot
\max ( \Vert L_0 ^{-1} \Vert, \Vert U_0 ^{-1} \Vert) \cdot{} \\
\hspace{3.5cm} \max \big( \kappa(L_0) \Vert U_0^{-1} \Vert, 
  \kappa(U_0) \Vert L_0^{-1} \Vert \big)^2.
\end{array}$$

\noindent
{\bf Numerical experiments.}
\todo{Do it.}

@@

\noindent
{\bf Charts.}
Let $V$ and $V^\c$ be two complementary subspaces of $E$ 
(\emph{i.e.} $V \oplus V^\c = E$). We assume that $V$ has 
dimension $d$ and denote by $\pi$ the projection $E \to V$ 
corresponding to the above decomposition. We introduce the set 
$\mathcal U_{V,V^\c}$ of all embeddings $f : V \hookrightarrow E$ 
such that $\pi \circ f = \id_V$. Clearly it is an affine space over
$\Hom(V,V^\c)$. 
Furthermore, we can embed it into $\Grass(E,d)$ by taking $f$ as
above to its image. This way, $\mathcal U_{V,V^\c}$ appears as
an open subset of $\Grass(E,d)$ consistant exactly of those subspaces 
$W$ such that $W \cap V^\c = 0$. As a consequence, the tangent space 
at each such $W$ becomes isomorphic to $\Hom(V,V^\c)$. The
identification $\Hom(V,V^\c) \to \Hom(W, E/W)$ is given by
$du \mapsto (du \circ f^{-1}) \text{ mod } W$ where $f : V 
\stackrel{\sim}{\to} W$ is the linear mapping defining $W$.

When the couple $(V, V^\c)$ varies, the open subsets $\mathcal 
U_{V,V^\c}$ cover the whole grassmaniann and then define an atlas 
of it. When implementing vector spaces on computer, we usually restrict 
ourselves to the subatlas consisting of all charts of the form $(V_I, 
V_{I^\c})$ where $I$ runs over the family of subsets of $\{1, 
\ldots, n\}$ of cardinality $d$ and $V_I$ is the subspaces spanned by 
the $e_i$'s with $i \in I$. A subspace $W \in E$ then belongs to at 
least one $\mathcal U_{V_I, V_{I^{\text{c}}}}$ and, given a family of
generators of $W$, we can determine such an $I$ together with the
corresponding embedding $f : V_I \hookrightarrow E$ by echelonizing 
the matrix of generators of $W$.

@@

\noindent
{\bf Direct images.}
Let $E$ and $F$ be two finite dimensional $K$-vector spaces of dimension 
$n$ and $m$ respectively. Let $d$ be an integer in $[0,n]$. We are
interested here in the function $\DI$ (for ``direct image'') defined 
over $\mathcal M = \Hom(E,F) \times \Grass(E,d)$ that takes the 
couple $(f,V)$ to $f(V)$. Since the dimension of $f(V)$ may \emph{a 
priori} vary, the map $\DI$ does not take its values in well-defined 
grassmannian. To tackle this problem, we stratify $\mathcal M$ as
follows: for each integer $r \in [0,d]$, we introduce the subset
$\mathcal M_r \subset \mathcal M$ consisting of those couples $(f,V)$
for which $f(V)$ has dimension $r$. The $\mathcal M_r$'s are
locally closed in $\mathcal M$ and therefore are submanifolds. 
Moreover, $\DI$ induces by restriction respectable differentiable 
functions between $K$-manifolds
$\DI_r : \mathcal M_r \to \Grass(F,r)$.

>>>>

\newtheorem{theo}{Theorem}[section]
\newtheorem{lem}[theo]{Lemma}
\newtheorem{prop}[theo]{Proposition}
\newtheorem{cor}[theo]{Corollary}
\newtheorem{quest}[theo]{Question}
\newtheorem{conj}[theo]{Conjecture}
%\theoremstyle{definition}
\newtheorem{rem}[theo]{Remark}
\newtheorem{ex}[theo]{Example}
\newtheorem{deftn}[theo]{Definition}
\newtheorem{rmk}[theo]{Remark}

@@

%\newcommand{\detp}{\text{\rm det}'}

%\DeclareMathOperator{\det}{det}
\newcommand{\detp}{\det'}
\newcommand{\DI}{\text{\rm DI}}
\newcommand{\II}{\text{\rm II}}
\DeclareMathOperator{\charpoly}{char}
\newcommand{\charp}{\charpoly'}

@@

\numberofauthors{3}
\author{
\alignauthor Xavier Caruso\\
  \affaddr{Universit\'e Rennes 1}\\
  \email{\normalsize \textsf{xavier.caruso@normalesup.org}}
\alignauthor Tristan Vaccon\\
  \affaddr{Universit\'e Rennes 1}\\
  \email{\normalsize \textsf{tristan.vaccon@univ-rennes1.fr}}
\alignauthor David Roe \\
  \affaddr{University of British Columbia}\\
  \email{\normalsize \textsf{roed.math@gmail.com}}
}

@@

\begin{abstract}
Using the differential precision methods developed previously by the same authors,
we study the $p$-adic stability of standard operations on matrices and vector 
spaces. We demonstrate that lattice-based methods surpass naive methods in many
applications, such as matrix multiplication and sums and intersections of subspaces.
We also analyze determinants and characteristic polynomials using these differential methods.
We supplement these observations with numerical experiments.
\end{abstract}

@@

To make this sentence more explcit, let us consider be a finite 
dimensional\footnote{The framework of \cite{caruso-roe-vaccon:14a} is 
actually those of Banach spaces. However we will not need infinite 
dimensional spaces in the present paper.} normed vector space $E$ 
defined over $K$. We use the notation $\Vert \cdot \Vert_E$ for the norm 
on $E$ and $B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. 
closed) ball of radius $r$ centered at the origin. A lattice $L \subset 
E$ is, by definition, a sub-$\O_K$-module which generates $E$ over $K$. 
Since we are working in a ultrametric world, the balls $B^{\phantom 
-}_E(r)$ and $B^-_E(r)$ are examples of lattices. Actually, lattices 
should be thought as special neighborhoods of $0$ and therefore are good 
candidates to model precision data. Moreover, as revealed in 
\cite{caruso-roe-vaccon:14a}, they behave quite well under (strictly) 
differentiable maps:

@@

\noindent
{\bf About the choice of the function $h$}.
At the beginning of \S \ref{ssec:boundLambdaf}, we have introduced a 
function $h$ in our differential equation. The aim of this was to be 
more flexible and possibly get this way better bounds on $\Lambda(f)$. 
In this paragraph, we discuss a recommended choice for it.

Actually, we do recommend taking $h$ as the identity function 
\emph{but} changing the norm on the codomain of $h$ at the same time.
More precisely, if $F$ is equipped with the norm $\Vert \cdot \Vert_F$, 
we recommand endowing it with the second norm defined by
$\Vert x \Vert'_F = \lambda \cdot \Vert x \Vert_F$ ($x \in F)$
where $\lambda$ is a positive real number that we shall choose later 
and taking $h : (F, \Vert \cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ 
acting as the identity on the underlying vector spaces. Doing this, the 
function $\Lambda(h)$ maps $x$ to $x + \log \lambda$ ($x \in \R$) and we 
choose $\Lambda_h = \Lambda(h)$.

We believe that a good way to optimize $\lambda$ is to choose it so that 
the two quantities in the minima in Proposition \ref{prop:boundLambdaf2} 
agree. With the expression we got just above for $\Lambda_h$, our 
heuristic yields $\log \lambda = \mu$, \emph{i.e.} $\lambda = e^\mu$.
Making this choice, the norms of $f(x)$ and $h(x)$ are comparable --- 
at least, on the domain we are interested in --- and we do not loose so 
much when we write that the norm of the couple $(f(x), h(x))$ is the 
maximum of those of $f(x)$ and $h(x)$.

@@

\noindent
{\bf Smith form.}
In what follows, we will often use the \emph{Smith decomposition} of a 
matrix $M \in M_{n,m}(K)$: we recall that it is a factorization of $M$ 
as a product $U_M \cdot \Delta_M \cdot V_M$ when $U_M$ and $V_M$ are 
unimodular (square) matrices of size $n$ and $m$ respectively, with 
coefficients in $\O_K$ and $\Delta_M$ satisfies the two following
conditions:

\noindent
(i) its $(i,j)$-th entry vanishes as soon as $i \neq j$

\noindent
(ii) its $(i,i)$-th entry divides its $(j,j)$-th entry when $i
\leq j$.

\noindent
Such a factorization is \emph{not} unique but the valuations of
the ``diagonal'' entries of $\Delta_M$ are. Smith decompositions
can be easily computed by performing elementary operations on rows and columns.
Regarding to precision, one can prove that if the entries of $M$ are all 
known up to some $O(\pi^N)$, it is possible to compute a factorization 
$U_M \cdot \Delta_M \cdot V_M$ where $U_M$ and $V_M$ are \emph{exact}
matrices satisfying the above conditions and where $\Delta_M$ is known
at precision $O(\pi^N)$ and satisfies~(i) and~(ii) modulo $\pi^N$.

@@

We now come to the computation of the determinant of the square $n 
\times n$ with coefficients in $K$. The differential computation is 
quite classical.  The differential of the function $\det : M_{n,n}(K) \to 
K$ at some point $M$ is the linear mapping
$$\detp(M) : dM \mapsto \tr(\com(M) \cdot dM),$$
where $\com(M)$ stands for the comatrix of $M$, which is $\det(M) M^{-1}$ when $M$ is invertible. If $M$ is not 
invertible, $\detp(M)$ is not surjective and we cannot apply Proposition 
\ref{prop:precision}. Therefore, we suppose that $M$ is invertible for the rest of this section.

@@

In terms of precision, Proposition \ref{prop:detmatrix} implies that if 
$M$ is given at flat precision $O(\pi^N)$ with $N > v$, then $\det(M)$ 
is known at precision $O(\pi^{N+v})$, so we are gaining $v$ significant 
digits in absolute precision\footnote{We note that the valuation of 
$\det(M)$ is $v + m_n$; thus if we are considering \emph{relative} 
precisions then we should say that we are loosing $m_n$ significant 
digits.}.
Furthermore, there exists a easy method for computing the determinant
of $M$ with this optimal precision. It consists in first computing an
approximate Smith decomposition of $M$:
$$M = \tilde U_M \cdot \tilde \Delta_M \cdot \tilde V_M.$$
Here $\tilde U_M$ and $\tilde V_M$ are exact unimodular matrices and
$\tilde \Delta_M$ is known at precision $O(\pi^N)$, is diagonal modulo
$\pi^N$ and its diagonal entries $a_1, \ldots, a_m$ have valuations 
$m_1, \ldots, m_n$ respectively.
We then have $\det(M) = \det(\tilde \Delta_M)$ and factoring $a_i$ on
the $i$-th row of $\tilde \Delta_M$, we obtain
$\det(\tilde \Delta_M) = a_1 \cdots a_m \cdot U$
where $U$ is a matrix which is congruent to the identity modulo 
$\pi^{N-m_n}$. Therefore we get:
$$\det(M) \equiv a_1 \cdots a_m \pmod{\pi^{N+v}}$$
from what we see that we can compute $\det(M)$ at the required precision
by simply evaluating the product $a_1 \cdots a_m$.

@@

Write $\charpoly : M_{n,n}(K) \to K[X]$ for the characteristic polynomial, and $K_{<n}[X] \subset K[X]$ for the subspace consisting of polynomials of degree less than $n$.  Then the determinant of $\charpoly$ at a point $M$ is
\[
\charp(M) : dM \mapsto \tr(\com(X - M) \cdot dM).
\]
The image is the $K$-span of the entries of $\com(X - M)$, which is clearly contained within $K_{<n}[X]$.  In fact, the image will equal $K_{<n}[X]$ as long as $M$ does not have two Jordan blocks with the same generalized eigenvalue.

For any precision lattice $H$, the precision of the characteristic polynomial will thus be the lattice in $K_{<n}[X]$ generated by the polynomials $\tr(\com(X - M) \cdot dM)$, as $dM$ ranges over a set of generators for $H$.  In the standard case that the precision lattice for $M$ is just $H = \pi^k M_{n,n}(\O_K)$, we can further analyze $\charp(H)$.

Recall that the \emph{Newton polygon} of a polynomial $\sum_{i} a_i X^i$ is the lower convex hull of the points $(i, \val(a_i))$ and the \emph{Newton polygon} of a matrix is the Newton polygon of its characteristic polynomial.  If $m_1 \le \cdots \le m_n$ are the valuations of the diagonal entries of $\Delta_M$ in a Smith factorization $U_M \cdot \Delta_M \cdot V_M$ of $M$ as in the previous section, the \emph{Hodge polygon} of $M$ is the lower convex hull of the points $(i, \sum_{j = i+1}^n m_i)$.

\begin{deftn}
The \emph{precision polygon} of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X - M)$.  
\end{deftn}

Any such polygon $P$ naturally yields a lattice in $K_{<n}[X]$, consisting of polynomials whose Newton polygons lie above $P$.  The lattice associated to $P$ is generated by monomials $a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.

If a polygon $P$ has vertices $(x_i, y_i)$, we $T_n(P)$ to be the translated polygon with vertices $(x_i - n, y_i)$ and $S_n(P)$ be the translated polygon with vertices $(x_i, y_i + n)$.

\begin{prop}
Suppose a matrix $M$ is known with precision $H = \pi^k M_{n,n}(\O_K)$.  Then the precision of the characteristic polynomial of $M$ is contained within $S_k$ applied to the lattice of the precision polygon of $M$.
\end{prop}
\begin{proof}
The Newton polygon of each entry of $\com(X - M)$ lies above the precision polygon by definition.  Since these generate $\charp(M)(H)$, the result follows.
\end{proof}

\begin{prop} \label{prop:prec_hodge}
The precision polygon of $M$ lies above $T_1$ applied to the Hodge polygon of $M$.
\end{prop}

\begin{proof}
\todo{Add this proof from Xavier's e-mail}
\end{proof}

For many matrices, the precision polygon equals the Hodge polygon.  For integer matrices with entries uniformly chosen between $0$ and $1024$, the $2$-adic precision polygon is equal to the Hodge polygon in $99.5\%$ of cases in dimension $4$, down to $99.1\%$ in dimension $8$.  \todo{p=3?}  Moreover, experiments support the following conjectures.

\begin{conj}
\begin{enumerate}
\item The precision polygon is equal to the Hodge polygon if the determinant of $M$ has valuation $0$ or $1$, or if $M$ is $3 \times 3$.
\item The height at $0$ of the precision polygon is equal to the height at $1$ of the Hodge polygon.
\item The precision polygon lies below $T_1$ applied to the Newton polygon.
\end{enumerate}
\end{conj}
Empirically, the precision polynomial seems most likely to differ from the Hodge polygon at $1$, corresponding to the precision of the linear term of the characteristic polynomial.

Of course, the precision polygon does necessarily agree with the full precision of the characteristic polynomial of $M$.  In $11\%$ of cases in dimension $3$, up to $15\%$ of cases in dimension $8$, the lattice associated to the precision polygon contains the actual precision lattice with index greater than $1$.  This percentage increases as the valuation of the determinant increases, reaching $34\%$ in dimension $9$ for matrices constrained to have determinant with $2$-adic valuation $12$.

Moreover, one can construct examples with arbitrarily large extra lattice precision.  If $M$ has elementary divisors with valuations $m_1 \le \cdots \le m_n$, Proposition \ref{prop:prec_hodge} implies that the precision lattice of the characteristic polynomial of $M$ is contained within $\O_{K,n}[X]$ with index at least $\sum_{i=1}^{n-1} m_i$.  The precision lattice of $1 + M$ is obtained from the precision lattice of $M$ via the transformation $X \mapsto X + 1$, but the precision polygon is now flat with height $0$.

@@

\todo{Improve this Subsection.}

In this Subsection, we apply our framework to the case of $LU$ factorization. The computation of the differential of this factorization has already been done in \cite[Appendix B]{caruso-roe-vaccon:14a}. It remains to determine under which condition Proposition \ref{prop:precision} applies.

\begin{deftn}
Let $n \in \mathbb{Z}_{\geq 0}$ and  $O_n \subset K^{n \times n}$ be the set of the matrices such that all principal minors are non-vanishing. Let $U_n$ (resp. $L_n$)  be the $K$-vector spaces of the upper-triangular (resp. the strictly lower-triangular) matrices in $K^{n \times n}.$ Let $L_n^u \subset K^{n \times n} $ be the set of lower-triangular matrices with $1$'s one the diagonal. We define $D : O_n \rightarrow L_n^u \times U_n$ to be the mapping sending a matrix $M$ to $(L,M)$ such that $LU$ is the LU factorization of $M.$
\end{deftn}

This example has been previously studied by X.Caruso in \cite{caruso:12a}, but without using differential precision.
We show here that the framework of Subsection \ref{ssec:boundLambdaf} applies directly.

\subsubsection*{Computation of the differential of $D$} 
We begin with some notations:
\begin{deftn}
We denote by $\textrm{low}$ (resp. $\textrm{up}$) the canonical projection $K^{n \times n} \rightarrow L_n$ (resp. $K^{n \times n} \rightarrow U_n$).
\end{deftn}

Then, an explicit formula for $D'$ has been computed in \cite[Appendix B]{caruso-roe-vaccon:14a} :
\begin{prop}
Let $(M,L,U) \in O_n \times L_n^u \times U_n$ be such that $M=LU.$ Then,

\begin{tabular}{cccc}
$D'(M) :$ &  $K^{n \times n}$ & $\rightarrow$ & $L_n \times U_n$ \\ 
 & $\mathrm{d}M$ & $ \mapsto$ & $(L \textrm{low}(L^{*-1} \mathrm{d}M U^{-1}), \textrm{up}(L^{*-1} \mathrm{dM} U^{-1}) U).$ \\ 
\end{tabular} 
\end{prop}

A right inverse of $D'(M)$ can be made explicit:
\begin{rem} \label{rem:rightinverseLU}
Let \begin{tabular}{cccc}
$E(M) :$ & $L_n \times U_n$  & $\rightarrow$ &  $K^{n \times n}$\\ 
 & $(A,B)$ & $ \mapsto$ & $AU+LB$. \\ 
\end{tabular}

Then $D'(M)  \circ E(M) = Id.$
\end{rem}

\subsubsection*{Applying Subsection \ref{ssec:boundLambdaf}}

In order to apply Subsection \ref{ssec:boundLambdaf}, $f$ will be a translation of $D,$ so as to ensure $g(0) \neq 0.$ Namely,  let $(M_0,L_{0},U_{0}) \in O_n \times L_n^u \times U_n$ be such that $M_0 = L_{0}U_{0},$ we define

\begin{tabular}{cccc}
$f :$ &  $K^{n \times n}$ & $\rightarrow$ & $L_n \times U_n$ \\ 
 & $M$ & $ \mapsto$ & $(L_{M+M_0}-L_{0}, U_{M+M_0}-U_{0}).$ \\ 
\end{tabular}

Then, $f(0)=0$ and \[f'(M) \cdot \mathrm{d}M = (L_{M+M_0} \textrm{low}(L_{M+M_0}^{-1} \mathrm{d}M U_{M+M_0}^{-1}), \textrm{up}(L_{M+M_0}^{-1} \mathrm{dM} U_{M+M_0}^{-1}) U).\]

We now proceed with the three steps of the framework of Subsection \ref{ssec:boundLambdaf}.

\textbf{The $C$ constant:}

\begin{lem}
For $C = \max (\Vert U_{0} \Vert,\Vert L_{0} \Vert ),$ $B_{K^{n \times n}}(1) \subset f'(0) \cdot B_{K^{n \times n}}(C).$ 
\end{lem}
\begin{proof}
Apply Remark \ref{rem:rightinverseLU}.
\end{proof}

\textbf{The differential equation:}

\begin{lem} Let $g=g_{M_0}$ be defined by:
\begin{align*}
g(A,B) \cdot X =& ( (L_0^*+A) \textrm{low} \left( (L_0^*+A)^{-1}X(U_0+B)^{-1} \right), \\
& \: \textrm{up} \left( (L_0^*+A)^{-1}X(U_0+B)^{-1} \right) (U_0+B) ).\\
\end{align*}
Then, \[f' = g \circ f.\]
\end{lem}

\textbf{Determining $\mu$ and $\nu$}

Let $\vert \: \vert $ be an ultrametric norm on $K$ and $\Vert \: \Vert$ be the corresponding subordinate matrix norm. To determine $\mu$ and $\nu,$ we first use the following lemma

\begin{lem}
Let $(U,V) \in K^{n \times n} \times K^{n \times n}.$ If $\Vert V \Vert < \Vert U \Vert$ then $\Vert U+V \Vert = \Vert U \Vert.$

If $U \in GL_n(K)$ and $\Vert U \Vert < \Vert U^{-1} \Vert^{-1},$ then $\Vert (U+V)^{-1} \Vert = \Vert U^{-1} \Vert.$

Finally $\Vert U^{-1} \Vert^{-1} \leq \Vert U \Vert.$
\end{lem}

This leads to the following upper-bound on the norm of $g(A,B)$.

\begin{prop}
Let $(A,B) \in L_n \times U_n.$ be such that $\Vert A \Vert < \Vert L_0^{*-1} \Vert $ and $\Vert B \Vert <\Vert U_0^{*-1} \Vert. $ Then 
\[ \Vert g(A,B) \cdot X \Vert \leq \max \left( \Vert L_0^* \Vert \Vert L_0^{*-1} \Vert  \Vert U_0^{*-1} \Vert, \Vert U_0^* \Vert \Vert U_0^{*-1} \Vert  \Vert L_0^{*-1} \Vert \right) \times \Vert X \Vert . \]
\end{prop}

\begin{prop}
For $f$ and $g$ as before, we can take $\nu = \min ( \Vert L_0^{-1} \Vert ^{-1} ,\Vert U_0^{-1} \Vert ^{-1} ),$ and \[ \mu = \max \left( \Vert L_0^* \Vert \Vert L_0^{*-1} \Vert  \Vert U_0^{*-1} \Vert, \Vert U_0^* \Vert \Vert U_0^{*-1} \Vert  \Vert L_0^{*-1} \Vert \right).\]
\end{prop}

\textbf{Conclusion}

Consequently, we can apply Proposition \ref{prop:boundLambdaf2} and Proposition 3.12 of \cite{caruso-roe-vaccon:14a}:

\begin{prop}
For any $\rho \in (0,1],$ if $r<e^{ \nu -2(\alpha +\mu)+\log \left( \frac{\rho}{C} \right)},$ the equation \ref{eq:firstorder} for $f$ is satisfied.
\end{prop}

\begin{ex}
If $\Vert L_0\Vert = \Vert L_0^{-1}\Vert =\Vert U_0\Vert =  \Vert U_0^{-1}\Vert = \rho=1$ and $p=2,$ then Proposition \ref{prop:precision} can be applied with $r=2^{-4}.$
\end{ex}

\subsubsection*{Numerical experiments}

\todo{Add numerical experiments...}

@@

\noindent
{\bf Charts.}
Let $V$ and $V^\c$ be two complementary subspaces of $E$ 
(\emph{i.e.} $V \oplus V^\c = E$). We assume that $V$ has 
dimension $d$ and denote by $\pi$ the projection $E \to V$ 
corresponding to the above decomposition. We introduce the set 
$\mathcal U_{V,V^\c}$ of all embeddings $f : V \hookrightarrow E$ 
such that $\pi \circ f = \id_V$. Clearly it is an affine space over
$\Hom(V,V^\c)$. 
Furthermore, we can embed it into $\Grass(E,d)$ by taking $f$ as
above to its image. This way, $\mathcal U_{V,V^\c}$ appears as
on open subset of $\Grass(E,d)$ consistant exactly of those subspaces 
$W$ such that $W \cap V^\c = 0$. As a consequence, the tangent space 
at each such $W$ becomes isomorphic to $\Hom(V,V^\c)$. The
identification $\Hom(V,V^\c) \to \Hom(W, E/W)$ is given by
$du \mapsto (du \circ f^{-1}) \text{ mod } W$ where $f : V 
\stackrel{\sim}{\to} W$ is the linear mapping defining $W$.

When the couple $(V, V^\c)$ varies, the open subsets $\mathcal 
U_{V,V^\c}$ cover the whole grassmaniann and then define an atlas 
of it. When implementing vector spaces on computes, we usually restrict 
ourselves to the subatlas consisting of all charts of the form $(V_I, 
V_{I^\c})$ where $I$ runs over the family of subsets of $\{1, 
\ldots, n\}$ of cardinality $d$ and $V_I$ is the subspaces spanned by 
the $e_i$'s with $i \in I$. A subspace $W \in E$ then belongs to at 
least one $\mathcal U_{V_I, V_{I^{\text{c}}}}$ and, given a family of
generators of $W$, we can determine such an $I$ together with the
corresponding embedding $f : V_I \hookrightarrow E$ by echelonizing 
the matrix of generators of $W$.

@@

\noindent
{\bf Direct images.}
Let $E$ and $F$ be two finite dimensional $K$-vector spaces of dimension 
$n$ and $m$ respectively. Let $d$ be an integer in $[0,n]$. We are
interested here in the function $\DI$ (for ``direct image'') defined 
over $\mathcal M = \Hom(E,F) \times \Grass(E,d)$ that takes the 
couple $(f,V)$ to $f(V)$. Since the dimension of $f(V)$ may \emph{a 
priori} vary, the map $\DI$ does not take its values in well-defined 
grassmannian. To tackle this problem, we stratify $\mathcal M$ as
follows: for each integer $r \in [0,d]$, we introduce the subset
$\mathcal M_r \subset \mathcal M$ consisting of those couples $(f,V)$
for which $f(V)$ has dimension $r$. The $\mathcal M_r$'s are
locally closed in $\mathcal M$ are therefore are submanifolds. 
Moreover, $\DI$ induces by restriction respectable differentiable 
functions between $K$-manifolds
$\DI_r : \mathcal M_r \to \Grass(F,r)$.
